---
title: "Microbiota adaptation review - 2. Analysis"
author:
- Garazi Martin Bideguren, garazi.bideguren@sund.ku.dk
- Antton Alberdi, antton.alberdi@sund.ku.dk
date: "07-11-2023"
output:
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data preparation

```{r load_libraries, echo=TRUE, message=FALSE, results=FALSE }
library(knitr)
library(dplyr)
library(gridExtra)
library(tidyverse)
library(devtools)
library(ggplot2)
library(vegan)
library(FSA)
```

Prepare the main dataset were all the analysed studies will have their obtained scores for each of the criteria. Subsample the dataset in 4 dataframes: 1)Select the columns corresponding to the 10 analysed criteria, 2)Select the columns corresponding to the experimental design, 3)Select the columns corresponding to the methodological resolution and 4) Select the columns corresponding to the reproducibility.
Finally, prepare the file with the weigthed values for each of the analysed criteria.

```{r load_data}
# Dataset
scores <- read.csv("data/scores.csv",header=T)

# Groups of criteria
criteria <- colnames(scores)[c(4:13)]
design <- colnames(scores)[c(4:7)]
methods <- colnames(scores)[c(8:12)]
reproducibility <- colnames(scores)[13]

# Weights
weights <- read.csv("data/weights.csv",header=T,row.names=1) %>%
   rowwise() %>%
   mutate(average=mean(c_across(everything()))) #calculate average weight
```

# Performance score statistics
Compute total performance scores given the vector of average criterion weights.

```{r overall_scores}
# Generate weight vectors
weight_consensus <- weights$average
names(weight_consensus) <- criteria
weight_consensus_design <- weight_consensus[design]/sum(weight_consensus[design]) # weighted criteria only for design
weight_consensus_methods <- weight_consensus[methods]/sum(weight_consensus[methods]) # weighted criteria only for methods
weight_consensus_reproducibility <- weight_consensus[reproducibility]/sum(weight_consensus[reproducibility]) # weighted criteria only for reproducibility

# Calculate weighted averages
scores <- scores %>%
  mutate(total_design = rowSums(across(all_of(design), ~ . * weight_consensus_design[[(cur_column())]]))) %>%
  mutate(total_methods = rowSums(across(all_of(methods), ~ . * weight_consensus_methods[[(cur_column())]]))) %>%
  mutate(total_reproducibility = rowSums(across(all_of(reproducibility), ~ . * weight_consensus_reproducibility[[(cur_column())]]))) %>%
  mutate(total = rowSums(across(all_of(criteria), ~ . * weight_consensus[[(cur_column())]])))
```

## Overall visual statistics
#### Studies per total score
Distribution of total scores of the 109 papers analysed
```{r histogram_score, echo=TRUE}
scores %>%
  ggplot(aes(x=total)) +
  geom_histogram(binwidth = 0.05,) +
  theme_classic()
```

#### Studies per year
The 109 papers analysed distributed per year
```{r histogram_year, echo=TRUE}
scores %>%
  ggplot(aes(x=Year)) +
  geom_histogram(binwidth = 1) +
  theme_classic()
```

# Performance scores per continent (Fig. S2)

```{r separate_continent, echo=TRUE}
scores_continent <- scores %>%
  separate_rows(Continent, sep = ", ") 
```

```{r boxplot_continent, eval=FALSE}
scores %>%
  mutate(Year = factor(Year)) %>%
  separate_rows(Continent, sep = ", ") %>%
  ggplot(aes(x = Continent, y = Total, fill = Continent)) +
  geom_violin() +
  geom_point(stat = "summary", fun = "median", shape=8, color = "black", size = 3) +
  geom_jitter(
    width = 0.15,
    height = 0,
    alpha = 0.5,
    size = 1.5) +
  scale_fill_manual(values = c("#E69F00", "#AA3377", "#228833", "#66CCEE", "#CCBB44", "#EE6677", "#4477AA")) +
  scale_colour_manual(values = c("#E69F00", "#AA3377", "#228833", "#66CCEE", "#CCBB44", "#EE6677", "#4477AA"))+
  theme_classic() +
  labs(title = "Total Score Distributions by Continent",
       y = "Total Score",
       x = "Continent") +
  coord_flip()
```

### Statistical model (Table S2)

```{r total_performance_continent_model, echo=TRUE}

# Kruskal-Wallis test
kruskal <- scores_continent %>%
  kruskal.test(total ~ Continent, .)

# Render results
kable(c(kruskal$statistic,kruskal$parameter,pvalue=kruskal$p.value))

# Dunn test for multiple comparisons
dunn <- scores_continent %>%
  dunnTest(total ~ Continent, data=., method="bh")

# Render results
kable(dunn$res)
```

# Performance scores per animal taxa

```{r separate_taxa, echo=TRUE}
scores_taxa <- scores %>%
  separate_rows(Taxa, sep = ", ") %>%
  mutate(Taxa = ifelse(Taxa == "Mammalia", "Mammals", Taxa)) %>%
  mutate(Taxa = ifelse(Taxa == "Aves", "Birds", Taxa)) %>%
  mutate(Taxa = ifelse(Taxa == "Reptilia", "Reptiles", Taxa)) %>%
  mutate(Taxa = ifelse(Taxa == "Amphibia", "Amphibians", Taxa)) %>%
  mutate(Taxa = ifelse(Taxa == "Dipnoi", "Fish", ifelse(Taxa == "Actinopterygii", "Fish", Taxa))) %>%
  mutate(Taxa = factor(Taxa, levels = c("Mammals","Birds","Reptiles","Amphibians","Fish")))
```

### Visualisation (Fig. 2c)

```{r boxplot_class, echo=TRUE, warning=FALSE}
scores_taxa %>%
  ggplot(aes(x = Taxa, y = total, fill = Taxa)) +
  geom_violin(color = NA) +
  geom_point(stat = "summary", fun = "median", shape=8, color = "black", size = 3) +
  geom_jitter(
    width = 0.15,
    height = 0,
    alpha = 0.5,
    size = 1.5) +
  scale_fill_manual(values = c("#AB67AA50", "#C66E6A50", "#CCC26A50", "#8FC96750", "#68C8C650")) +
  scale_colour_manual(values = c("#AB67AA", "#C66E6A", "#CCC26A", "#8FC967", "#68C8C6"))+
  theme_classic() +
  labs(y = "Performance Score") +
  coord_flip() +
  theme(legend.position = "none")
```


### Statistical model (Table S2)

```{r total_performance_taxa_model, echo=TRUE}

# Kruskal-Wallis test
kruskal <- scores_taxa %>%
  kruskal.test(total ~ Taxa, .)

# Render results
kable(c(kruskal$statistic,kruskal$parameter,pvalue=kruskal$p.value))

# Dunn test for multiple comparisons
dunn <- scores_taxa %>%
  dunnTest(total ~ Taxa, data=., method="bh")

# Render results
kable(dunn$res)
```


### Visualisation (Fig. S2) **Were do we place this in the paper??**

```{r performance_domains_taxa_plot, echo=TRUE}

# Experimental design
total_design_taxa_plot <- scores_taxa %>%
  ggplot(aes(x = Taxa, y = total_design, fill = Taxa)) +
  geom_violin(color = NA) +
  geom_point(stat = "summary", fun = "median", shape=8, color = "black", size = 3) +
  geom_jitter(
    width = 0.15,
    height = 0,
    alpha = 0.5,
    size = 1.5) +
  scale_fill_manual(values = c("#AB67AA50", "#C66E6A50", "#CCC26A50", "#8FC96750", "#68C8C650")) +
  scale_colour_manual(values = c("#AB67AA", "#C66E6A", "#CCC26A", "#8FC967", "#68C8C6"))+
  theme_classic() +
  labs(y = "Experimental design") +
  coord_flip() +
  theme(legend.position = "none", axis.title.y = element_blank())

# Methodological resolution
total_methods_taxa_plot <- scores_taxa %>%
  ggplot(aes(x = Taxa, y = total_methods, fill = Taxa)) +
  geom_violin(color = NA) +
  geom_point(stat = "summary", fun = "median", shape=8, color = "black", size = 3) +
  geom_jitter(
    width = 0.15,
    height = 0,
    alpha = 0.5,
    size = 1.5) +
  scale_fill_manual(values = c("#AB67AA50", "#C66E6A50", "#CCC26A50", "#8FC96750", "#68C8C650")) +
  scale_colour_manual(values = c("#AB67AA", "#C66E6A", "#CCC26A", "#8FC967", "#68C8C6"))+
  theme_classic() +
  labs(y = "Methodological resolution") +
  coord_flip() +
  theme(legend.position = "none", axis.title.y = element_blank())

# Reproducibility
total_reproducibility_taxa_plot <- scores_taxa %>%
  ggplot(aes(x = Taxa, y = total_reproducibility, fill = Taxa)) +
  geom_violin(color = NA) +
  geom_point(stat = "summary", fun = "median", shape=8, color = "black", size = 3) +
  geom_jitter(
    width = 0.15,
    height = 0,
    alpha = 0.5,
    size = 1.5) +
  scale_fill_manual(values = c("#AB67AA50", "#C66E6A50", "#CCC26A50", "#8FC96750", "#68C8C650")) +
  scale_colour_manual(values = c("#AB67AA", "#C66E6A", "#CCC26A", "#8FC967", "#68C8C6"))+
  theme_classic() +
  labs(y = "Reproducibility") +
  coord_flip() +
  theme(legend.position = "none", axis.title.y = element_blank())

#Composite plot
grid.arrange(grobs = list(total_design_taxa_plot,total_methods_taxa_plot,total_reproducibility_taxa_plot), layout_matrix = matrix(1:3, nrow = 1))
```

# Performance scores across years

Linear regression model between the performance score and the publication year.

### Statistical model (Table S1)

```{r total_performance_year_model, echo=TRUE}
total_performance_year_model <- lm(total ~ Year, data = scores) %>% summary()
total_performance_year_model_table <- total_performance_year_model$coefficients[2,c(1,3,4)] %>% round(.,4)

# Render table
kable(total_performance_year_model_table)
```

### Visualisation (Fig. 2b)
Linear regression between the performance score and the publication year.

```{r total_performance_year_plot, echo=TRUE}
scores %>%
  mutate(Year = as.numeric(Year)) %>%
  ggplot(aes(x=Year,y=total)) +
  geom_jitter(alpha=0.6, width=0.3) +
  ylim(0,1) +
  geom_smooth(method=lm, colour="#7A564A", fill = "#C1DDE040") +
  theme_classic() +
  labs(y = "Total performance", x = "Year")
```

## Temporal trend of performance domains

Linear regression between the performance domain scores and the publication year.

### Statistical model (Table S1)

```{r performance_domains_year_models, echo=TRUE}
# Experimental design
total_design_year_model <- lm(total_design ~ Year, data = scores) %>% summary()

# Methodological resolution
total_methods_year_model <- lm(total_methods ~ Year, data = scores) %>% summary()

# Reproducibility
total_reproducibility_year_model <- lm(total_reproducibility ~ Year, data = scores) %>% summary()

# Composite result table
performance_domains_year_models_table<- rbind(
      total_design_year_model$coefficients[2,c(1,3,4)],
      total_methods_year_model$coefficients[2,c(1,3,4)],
      total_reproducibility_year_model$coefficients[2,c(1,3,4)]
    )  %>% round(.,4)
rownames(performance_domains_year_models_table) <- c("Experimental design","Methodological resolution","Reproducibility")

# Render table
kable(performance_domains_year_models_table)
```

### Visualisation (Fig. 2a)
Total performance scores of each study with the relative contribution of each domain indicated by a different colour.

```{r stack_barplot, echo=TRUE}
scores %>%
  mutate(design = rowSums(across(all_of(design), ~ . * weight_consensus[[(cur_column())]]))) %>%
  mutate(methods = rowSums(across(all_of(methods), ~ . * weight_consensus[[(cur_column())]]))) %>%
  mutate(reproducibility = rowSums(across(all_of(reproducibility), ~ . * weight_consensus[[(cur_column())]]))) %>%
  select(DOI, Year, design, methods, reproducibility) %>%
  pivot_longer(cols = c(design, methods, reproducibility), names_to = "Category", values_to = "Score") %>%
  mutate(DOI = factor(DOI, levels = unique(DOI))) %>%
  mutate(Year = factor(Year, levels = sort(unique(Year)))) %>%
  ggplot(., aes(x = DOI, y = Score, fill = Category)) +
      geom_bar(stat = "identity") +
      labs(title = "Total scores",
           x = "Paper ID",
           y = "Score") +
      theme_minimal() +
      facet_grid(~ Year, scales = "free", space = "free")+
      theme(axis.text.x = element_blank(),axis.ticks.x=element_blank())+
      scale_fill_manual(values = c("design" = "#C1DDE0", "methods" = "#C0AF9D", "reproducibility" = "#7A564A")) +
      theme(legend.position = "none")
```

### Visualisation (Fig. S3)

```{r performance_domains_year_plot, echo=TRUE}
# Experimental design
total_design_year_plot <- scores %>%
  mutate(Year = as.numeric(Year)) %>%
  ggplot(aes(x=Year,y=total_design)) +
  geom_jitter(alpha=0.6, width=0.3) +
  ylim(0,1) +
  geom_smooth(method=lm, colour="#7A564A", fill = "#C1DDE040") +
  theme_classic() +
  labs(y = "Experimental design", x = "Year")

# Methodological resolution
total_methods_year_plot <- scores %>%
  mutate(Year = as.numeric(Year)) %>%
  ggplot(aes(x=Year,y=total_methods)) +
  geom_jitter(alpha=0.6, width=0.3) +
  ylim(0,1) +
  geom_smooth(method=lm, colour="#7A564A", fill = "#C1DDE040") +
  theme_classic() +
  labs(y = "Methodological resolution", x = "Year")

# Reproducibility
total_reproducibility_year_plot <- scores %>%
  mutate(Year = as.numeric(Year)) %>%
  ggplot(aes(x=Year,y=total_reproducibility)) +
  geom_jitter(alpha=0.6, width=0.3) +
  ylim(0,1) +
  geom_smooth(method=lm, colour="#7A564A", fill = "#C1DDE040") +
  theme_classic() +
  labs(y = "Reproducibility", x = "Year")

#Composite plot
grid.arrange(grobs = list(total_design_year_plot,total_methods_year_plot,total_reproducibility_year_plot), layout_matrix = matrix(1:3, nrow = 1))
```

# Performance scores vs. conclusiveness level

### Conclusiveness scores distribution (Fig. 2d)

```{r adaptation_means, echo=TRUE}
scores %>%
  select(DOI,Adaptation.means,Conclusiveness) %>%
  mutate(Conclusiveness = factor(Conclusiveness, levels = c(0,1,2,3,4))) %>%
  mutate(Adaptation.means = factor(Adaptation.means)) %>%
  #pivot_longer(cols = c(Adaptation.means,Conclusiveness), names_to = "Category", values_to = "Score") %>%
  ggplot(., aes(x = Adaptation.means, fill=Conclusiveness)) +
    geom_bar(position='stack', stat='count') +
    labs(x = "Adaptation as hypothesis?", y = "Number of papers") +
    scale_fill_manual(values = c("#6A697C", "#758A9B", "#C1DDE0", "#FFD367", "#F6A438")) +
    coord_flip() +
    theme_minimal() +
    theme(legend.position = "none")
```

### Statistical model 

```{r total_performance_conclusiveness_model, echo=TRUE}
total_performance_conclusiveness_model <- lm(total ~ Conclusiveness, data = scores) %>% summary()
total_performance_conclusiveness_model_table <- total_performance_conclusiveness_model$coefficients[2,c(1,3,4)] %>% round(.,4)

# Render table
kable(total_performance_conclusiveness_model_table)
```

### Visualisation (Fig. 2e)

```{r total_performance_conclusiveness_plot, echo=TRUE}
scores %>%
  mutate(Conclusiveness = as.numeric(Conclusiveness)) %>%
  ggplot(aes(x=Conclusiveness,y=total)) +
  geom_jitter(alpha=0.6, width=0.3) +
  ylim(0,1) +
  geom_smooth(method=lm, colour="#7A564A", fill = "#C1DDE040") +
  theme_classic() +
  labs(y = "Reproducibility", x = "Conclusiveness")
```
